Variance reduction for Markov chains via martingale representations

### 2020-04-30 aap/Revision/vr-mcmc-revision-new-new.tex
update

Now, from the H\"older inequality, we obtain (with $\|X\|_p = (\mathsf{E}X^p)^{\frac{1}{p}}$)
\begin{align*}
\mathsf{E}\left[\sum_{j=1}^{q}\prod_{k=l}^{j}\alpha_{k}\left( \sum_{p=j+1}^{q+1}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right] & \leq\sum_{j=1}^{q}\left\Vert \prod_{k=1}^{j}\alpha_{k}\right\Vert _{2}\left\Vert \sum_{p=j+1}^{q+1} \prod_{k=j+1}^{p}\alpha_{k}\right\Vert _{4}^{2}\\
 & \leq\sum_{j=1}^{q} \left\Vert \prod_{k=1}^{j}\alpha_{k}\right\Vert _{2}\left(\sum_{p=j+1}^{q+1}\left\Vert \prod_{k=j+1}^{p}\alpha_{k}\right\Vert _{4}\right)^{2}.
\end{align*}
Now using the fact that $\prod\limits_{k=j+1}^{p}\alpha_{k} \leq \exp\left(-\sum\limits_{k=j+1}^{p}\gamma\mu'(X_{k-1}^x)\right),$
we get
\[
\mathsf{E}\left[\sum_{j=1}^{q}\prod_{k=1}^{j}\alpha_{k}\left(\sum_{p=j+1}^{q+1}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right]\leq\sum_{j=1}^{q} \zeta^{1/2}_{1,j}(2) \left(\sum_{p=j+1}^{q+1} \zeta^{1/4}_{j+1,p}(4)\right)^{2},
\]
where we denote
\[
\zeta_{l,j}(u)=\mathsf{E}\left[\rme^{-u\gamma\sum_{k=l}^{j}\mu'(X^x_{k-1})}\right],\quad u>0.
\]
Note that $\mu^{\prime}(x)$ is bounded and Lipschitz due  our assumptions and from Lemma~\ref{lem:exp_bound} in Appendix~\ref{sec:appendix} it follows that
\begin{align*}
\zeta_{l,j}(u) \leq \begin{cases} C(x)\exp\left\{(j-l)\gamma\left(4u^2\kappa^{-2} \|f\|^2_{\text{Lip}} + M^2 + C_1u\sqrt{\gamma} -  u\pi(\mu')\right)\right\}, \quad \gamma(j-l) \geq C_0 := \left(\frac{4}{m+L} + \kappa^{-1}\right) \\
\exp\left\{uC_0\max\limits_{y}|\mu'(y)|\right\}, \quad 0 \leq \gamma(j-l) \leq C_0	
\end{cases}
\end{align*}
for constants $C(x)$ and $C_1$ which could be made precise. Furthermore, since $\mu\pi',\mu'\pi\in L^{1}(\mathbb{R})$ and $\pi^{\prime}(x) = -\frac{1}{2}\pi(x)\mu(x)$, we have
\[
\pi(\mu')=\int\mu'(x)\pi(x)\,dx=-\int\mu(x)\pi'(x)\,dx=\frac{1}{2}\int\mu^{2}(x)\pi(x)\,dx>0
\]
Hence the following bound follows
\begin{align*}
\zeta_{l,j}(u) \leq \begin{cases} C(x)\exp\left\{-\gamma(j-l)\alpha\right\}, \quad \gamma(j-l) \geq C_0 \\
\exp\left\{uC_0\max\limits_{y}|\mu'(y)|\right\}, \quad 0 \leq \gamma(j-l) \leq C_0	
\end{cases}
\end{align*}
with the quantity $ \alpha = \min(...,...) > 0$. Hence we obtain
\begin{align*}
\sum_{p=j+1}^{q+1} \zeta^{1/4}_{j+1,p}(4) &\leq \frac{C_0}{\gamma}\rme^{uC_0\max\limits_{y}|\mu'(y)|} + C(x)\sum_{p=j+1}^{\infty} \rme^{-\gamma(p-j)\alpha}
\leq \frac{C_0}{\gamma}\rme^{uC_0\max\limits_{y}|\mu'(y)|} + \frac{C(x)}{1-\rme^{- \gamma\alpha}}
\end{align*}
and
\begin{align*}
\sum_{j=1}^{q} \zeta^{1/2}_{1,j}(2) \leq \frac{C_0}{\gamma}\rme^{uC_0\max\limits_{y}|\mu'(y)|} + \frac{C(x)}{1-\rme^{- \gamma\alpha}}
\end{align*}
Thus the final bound follows:
\begin{align*}
\mathsf{Var}\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X^x_{p}\right)\right] \leq \gamma^{-1} C_K C(x)
\end{align*}
 with $C_K$  depending neither on $q$ nor on $\gamma$. The proof is completed.
\end{proof}

\appendix
\section{Bounds for moments of  ULA}\label{sec:appendix}

@@

\begin{lem}
\label{lem:v_ergodicity}
 Assume  {\bf (H1)} and {\bf (H2)}. Then for $0 < \gamma < \overline{\gamma}=m / 4L^2$, for any $x \in \rset^d$ it holds
$$
d_V(\delta_xQ_{\gamma}^{1,n},\pi_\gamma) \leq C\rho^n\left(V(x) + \pi_\gamma(V)\right)
$$
with $V(x) = 1 + \|x\|^2$ and constants
$$
C = \left(1 + \exp{\left(-\frac{m\gamma}{2}\right)}\right)\left(1+\frac{\overline{b}}{(1-\varepsilon)(1-\exp{\left(-\frac{m\gamma}{2}\right)} - \frac{2b}{1+d})}\right);
$$

$$
b = \gamma(\frac{25K_1^2}{8} + 2d + m); \quad \overline{b} = b\exp{\left(-\frac{m\gamma}{2}\right)} + d; \quad \varepsilon = 2\Phi\left(-\frac{\sqrt{d}(1+L\gamma)}{2\sqrt{\gamma}}\right); \quad
$$

$$
\rho = \exp{\left(-\gamma\left(\frac{m}{2} - 2\frac{\frac{25K_1^2}{8} + 2d + m}{d}\right)\frac{\log{(1-\varepsilon)}}{\log{(1-\varepsilon)} + \log{(\exp{\left(-\frac{m\gamma}{2}\right)} + \frac{2b}{d+1})}}\right)}
$$
\end{lem}

\begin{proof} Note that the condition {\bf (H1)} implies that the Markov kernel $Q^{1,n}_{\boldsymbol{\gamma}}$ satisfies $(1,\varepsilon)$-Doeblin condition with $\varepsilon = 2\Phi\left(-\frac{\sqrt{d}(1+L\gamma)}{2\sqrt{\gamma}}\right)$. Together with drift condition \eqref{eq:lyapunov} it allows to apply \cite[Theorem~19.4.1]{moulines2018} with appropriate constants.
\end{proof}

@@

 with $0 < \gamma < \frac{4}{m+L}, \quad \mu(x) = \mu_{0}(x) + \Delta(x)$, where $\mu_0^{\prime}(x) <-m$ for some $m > 0$, $\mu_0(x) \in Lip(L)$. Suppose that there exists $M > 0,$ such that $\left|\Delta(x)\right| \leq M$.

@@

\PE_x \rme^{u\gamma\sum\limits_{k=N+1}^{N+n}f(X_k)} \leq C\exp\left\{n\gamma\left(2u^2\kappa^{-2} \|f\|^2_{\text{Lip}}\left(1+\frac{1}{n\gamma}\left(\frac{4}{m+L} + \kappa^{-1}\right)\right) + M^2 + C_1u\sqrt{\gamma} +  u\pi(f)\right)\right\}

@@

with positive constants $C < \exp\left\{\frac{4u\|f\|_{\text{Lip}}}{mL}\int_{\rset}|x-y|\,\pi_{\gamma}(dy)\right\}$ and $C_1$ depending only on $m,L$.

>>>>

Now, from the H\"older inequality, we obtain (with $\|X\|_p = (\mathsf{E}X^p)^{\frac{1}{p}}$)
\begin{align*}
\mathsf{E}\left[\sum_{j=1}^{q}\prod_{k=l}^{j}\alpha_{k}\left( \sum_{p=j+1}^{q+1}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right] & \leq\sum_{j=1}^{q}\left\Vert \prod_{k=1}^{j}\alpha_{k}\right\Vert _{2}\left\Vert \sum_{p=j+1}^{q+1} \prod_{k=j+1}^{p}\alpha_{k}\right\Vert _{4}^{2}\\
 & \leq\sum_{j=1}^{q} \left\Vert \prod_{k=1}^{j}\alpha_{k}\right\Vert _{2}\left(\sum_{p=j+1}^{q+1}\left\Vert \prod_{k=j+1}^{p}\alpha_{k}\right\Vert _{4}\right)^{2}.
\end{align*}
Now using the fact that $\prod\limits_{k=j+1}^{p}\alpha_{k} \leq \exp\left(-\sum\limits_{k=j+1}^{p}\gamma\mu'(X_{k-1}^x)\right),$
we get
\[
\mathsf{E}\left[\sum_{j=1}^{q}\prod_{k=1}^{j}\alpha_{k}\left(\sum_{p=j+1}^{q+1}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right]\leq\sum_{j=1}^{q} \zeta^{1/2}_{1,j}(2) \left(\sum_{p=j+1}^{q+1} \zeta^{1/4}_{j+1,p}(4)\right)^{2},
\]
where we denote
\[
\zeta_{l,j}(u)=\mathsf{E}\left[\rme^{-u\gamma\sum_{k=l}^{j}\mu'(X^x_{k-1})}\right],\quad u>0.
\]
Note that due to our assumptios, $\mu(x) = \mu_0(x) + \Delta(x)$ with $\mu^{\prime}_0(x) < -m$. Thus by applying Holder inequality to $\zeta_{l,j}(u)$, we obtain
\begin{equation}
\label{eq:zeta}
\zeta_{l,j}(u) \leq \left(\mathsf{E}\left[\rme^{-2u\gamma\sum_{k=l}^{j}\mu_0'(X^x_{k-1})}\right]\right)^{1/2}\left(\mathsf{E}\left[\rme^{-2u\gamma\sum_{k=l}^{j}\Delta'(X^x_{k-1})}\right]\right)^{1/2}
\end{equation}
Note that $\left|\Delta^{\prime}(x)\right| \leq M^{\prime}$, hence
\[
\left(\mathsf{E}\left[\rme^{-2u\gamma\sum_{k=l}^{j}\Delta'(X^x_{k-1})}\right]\right)^{1/2} \leq \mathsf{E}\rme^{2u(j-l)M^{\prime}}
\]
To bound the first exponential moment in \ref{eq:zeta}, we apply
Lemma~\ref{lem:exp_bound} in Appendix~\ref{sec:appendix}. It implies that
\begin{align*}
\zeta_{l,j}(u) \leq \begin{cases} C(x)\exp\left\{(j-l)\gamma\left(4u^2\kappa^{-2} \|\mu^{\prime}\|^2_{\text{Lip}} + M^2 + M^{\prime} + C_1u\sqrt{\gamma} -  u\pi(\mu')\right)\right\}, \quad \gamma(j-l) \geq C_0 \\
\exp\left\{uC_0\max\limits_{y}|\mu'(y)|\right\}, \quad 0 \leq \gamma(j-l) \leq C_0	
\end{cases}
\end{align*}
with $C_0 = \left(\frac{4}{m+L} + \kappa^{-1}\right)$, and constants $C(x)$ and $C_1$ given in Lemma~\ref{lem:exp_bound} . Furthermore, since $\mu\pi',\mu'\pi\in L^{1}(\mathbb{R})$ and $\pi^{\prime}(x) = -\frac{1}{2}\pi(x)\mu(x)$, we have
\[
\pi(\mu')=\int\mu'(x)\pi(x)\,dx=-\int\mu(x)\pi'(x)\,dx=\frac{1}{2}\int\mu^{2}(x)\pi(x)\,dx>0
\]
Hence the following bound follows
\begin{align*}
\zeta_{l,j}(u) \leq \begin{cases} C(x)\exp\left\{-\gamma(j-l)\alpha\right\}, \quad \gamma(j-l) \geq C_0 \\
\exp\left\{uC_0\max\limits_{y}|\mu'(y)|\right\}, \quad 0 \leq \gamma(j-l) \leq C_0	
\end{cases}
\end{align*}
with the quantity $ \alpha = \min(...,...) > 0$. Hence we obtain
\begin{align*}
\sum_{p=j+1}^{q+1} \zeta^{1/4}_{j+1,p}(4) &\leq \frac{C_0}{\gamma}\rme^{uC_0\max\limits_{y}|\mu'(y)|} + C(x)\sum_{p=j+1}^{\infty} \rme^{-\gamma(p-j)\alpha}
\leq \frac{C_0}{\gamma}\rme^{uC_0\max\limits_{y}|\mu'(y)|} + \frac{C(x)}{1-\rme^{- \gamma\alpha}}
\end{align*}
and
\begin{align*}
\sum_{j=1}^{q} \zeta^{1/2}_{1,j}(2) \leq \frac{C_0}{\gamma}\rme^{uC_0\max\limits_{y}|\mu'(y)|} + \frac{C(x)}{1-\rme^{- \gamma\alpha}}
\end{align*}
Thus the final bound follows:
\begin{align*}
\mathsf{Var}\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X^x_{p}\right)\right] \leq \gamma^{-1} C_K C(x)
\end{align*}
 with $C_K$  depending neither on $q$ nor on $\gamma$. The proof is completed.
\end{proof}

\appendix
\section{Geometric ergodicity of the Unadjusted Langevin Algorithm}\label{sec:appendix}

@@

\begin{lem}
\label{lem:v_ergodicity}
 Assume  {\bf (H1)} and {\bf (H2)}. Then for $0 < \gamma < \overline{\gamma}= \min(m / 4L^2,1/L)$, for $x \in \rset^d: \|x\| \leq K$  it holds
$$
d_V(\delta_xQ_{\gamma}^{1,n},\pi_\gamma) \leq C\rho^n\left(V(x) + \pi_\gamma(V)\right)
$$
with $V(x) = 1 + \|x\|^2$ and constants
$$
C = \left(1 + \exp{\left(-\frac{m\gamma}{2}\right)}\right)\left(1+\frac{\overline{b}}{(1-\varepsilon)(1-\exp{\left(-\frac{m\gamma}{2}\right)} - \frac{2b}{2+K^2})}\right);
$$

$$
b = \gamma(\frac{25K_1^2}{8} + 2d + m); \quad \overline{b} = b\exp{\left(-\frac{m\gamma}{2}\right)} + K^2 + 1; \quad \varepsilon = 2\Phi\left(-K\sqrt{\left(1+\frac{1}{L}\right)\left(1 + 2L\right)}\right); \quad
$$

$$
\rho = \exp{\left(-\gamma\left(\frac{m}{2} - 2\frac{\frac{25K_1^2}{8} + 2d + m}{K^2 + 1}\right)\frac{\log{(1-\varepsilon)}}{\log{(1-\varepsilon)} + \log{(\exp{\left(-\frac{m\gamma}{2}\right)} + \frac{2b}{K^2+2})}}\right)}
$$
\end{lem}

\begin{proof} Note that the condition {\bf (H1)} implies that for any $x,y \in \rset^d$,
\[
\|x-y-\gamma(\nabla U(x) - \nabla U(y))\|^2 \leq (1 + 2L\gamma + \gamma^2)\|x-y\|^2
\] 
Hence, the result of [Corollary 5]\cite{DurmusBortoli} implies that
\[
\|\delta_xQ^{[1/\gamma]}_{\gamma} - \delta_yQ^{[1/\gamma]}_{\gamma}\|_{\text{TV}} \leq 2(1-\varepsilon)
\]
with $\varepsilon = 2\Phi\left(-K\sqrt{\left(1+\frac{1}{L}\right)\left(1 + 2L\right)}\right)$ for all $x,y$ such that $\|x\|,\|y\| \leq K$, which means that the kernel $Q_{\gamma}$ satisfies $([1/\gamma],\varepsilon)$-Doeblin codition on the set $\{x: V(x) < 1+K^2\}$. Together with drift condition \eqref{eq:lyapunov} it allows to apply \cite[Theorem~19.4.1]{moulines2018} with appropriate constants.
\end{proof}

@@

 with $0 < \gamma < 4(m+L)^{-1}, \quad \mu(x) = \mu_{0}(x) + \Delta(x)$, where $\mu_0^{\prime}(x) <-m$ for some $m > 0$, $\mu_0(x) \in Lip(L)$. Denote by 
 \pi(y) = \frac{\exp\left\{\int\limits_{0}^{y}\mu_{0}(x)\,dx\right\}}{\int\limits_{-\infty}^{+\infty}\exp\left\{\int\limits_{0}^{y}\mu_{0}(x)\,dx\right\}\,dy}
 the stationary distribution of the corresponding continuous-time diffusion.
 Suppose that there exists $M > 0,$ such that $\left|\Delta(x)\right| \leq M$.

@@

\PE_x \rme^{u\gamma\sum\limits_{k=N+1}^{N+n}f(X_k)} \leq C(x)\exp\left\{n\gamma\left(2u^2\kappa^{-2} \|f\|^2_{\text{Lip}}\left(1+\frac{1}{n\gamma}\left(\frac{4}{m+L} + \kappa^{-1}\right)\right) + M^2 + C_1u\sqrt{\gamma} +  u\pi(f)\right)\right\}

@@

with $\kappa = 2mL(m+L)^{-1}$, positive constant $C_1$ depending only on $m,L$ and 
C(x) = \exp\left\{\frac{4u\|f\|_{\text{Lip}}}{mL}\int_{\rset}|x-y|\,\pi_{\gamma}(dy)\right\}
